Factors affecting reliablity of state-space age-structured assessment models

### 2024-02-06 Project_0/paper/paper.tex

We used the Woods Hole Assessment Model (WHAM) to configure operating
and estimation models in our simulation study
\citep[]{millerstock20,stockmiller21}. WHAM is an R package freely
available as a github repository. For this study we used
\href{https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3}{version
1.0.6.9000, commit 77bbd94}. This packages has also been used to
configure operating and estimating models for closed loop simulations
evaluating index-based assessment methods \citep{legaultetal23} and is
used for management of haddock, butterfish, American plaice, bluefish,
Atlantic cod, and black sea bass in the Northeast US.

We completed a simulation study with a number of operating models that
can be categorized based on where process error random effects are
assumed: abundance at age (R, R+S), natural mortality (R+M), fleet
selectivity (R+Sel), or index catchability (R+q). For each operating
model assumptions about variance of process errors and observations are
required and the values we used were based on a review of the range of
estimates from recent applications of WHAM in management of stocks of
haddock, butterfish, and American plaice in the NE US.

In total, we configured 72 operating models with alternative assumptions
about the source and variability of process errors, level of observation
error in indices and age composition data, and contrast in fishing
pressure over time. We fitted 20 estimating models to observations from
each of 100 simulations where process errors were also simulated.
Estimating models made alternative assumptions about the source of
process errors and whether natural mortality (or the mean for models
with process error in natural mortality) was estimated and whether a
Beverton-Holt stock recruit relationship was estimated within the
estimation model. Details of each of the operating and estimating models
are described below.

We did not use the bias-correction feature for process errors or
observations described by \citet{stockmiller21} for operating and
estimating models.Simulations and model fitting were all carried out on
the University of Massachusetts Green High-Performance Computing
Cluster. All code we used to perform the simulation study and summarize
results can be found at
\url{https://github.com/timjmiller/SSRTWG/tree/main/Project_0/code}.

@@

The population model tracks 10 age classes: ages 1 to 10+ and we assume
spawning occurs 1/4 of the way through the year. The maturity at age was
a logistic curve with \(a_{50}\) = 2.89 and slope = 0.88 and assumed
known in all estimation models (Figure \ref{om_maturity}).

@@

We assumed a Beverton-Holt stock recruit function with constant
pre-recruit mortality parameters for all operating models. All
post-recruit productivity components are constant in the NAA and survey
catchability process error operating models. Therefore steepness and
unfished recruitment are also constant over the time period for those
operating models \citep{millerbrooks21}. We specified unfished
recruitment = \(R_0 = e^{10}\) and \(\Fmsy = F_{40\%} = 0.348\) equated
to a steepness of 0.69 and \(\alpha=0.60\) and
\(\beta = 2.4 \times 10^{-5}\) for the \[
N_{1,y} = \frac{\alpha \text{SSB}_{y-1}}{1 + \beta \text{SSB}_{y-1}} 
\] Beverton-Holt parameterization (Figure \ref{om_sr}). For OMs without
process errors on natural mortality we assumed the rate was assumed 0.2.
For OMs with process errors on natural mortality the mean rate was 0.2.

Two alternative fishing histories were used for operating models. In the
first scenario, the stock experiences overfishing for the first 20 years
and fishing at \Fmsy for the last 20 years. In the second scenario, the
stock is fished at \Fmsy for the entire time period. The magnitude of
the overfishing assumptions is based on average estimates of overfishing
for NE groundfish stocks from \citet{wiedenmannetal19}.
\Citet{legaultetal23} also used similar approaches to defining fishing
mortality histories for operating models.

@@

The factorial combinations of fishing histories, observation error
assumptions, and marginal variance and autocorrelation assumptions for
natural mortality results in 16 different operating models with random
effects on natural mortality (Table \ref{M_om_table}). Internally, WHAM
treats natural mortality as a log-transformed parameter \[
\log M_y = \mu_M + \epsilon_{M,y}
\] that is a linear combination of a mean \(\mu_M\) and any annual
random effects marginally distributed as
\(\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)\). The values
for the marginal variance and correlation parameters assumed in the
operating models are given in Table \ref{M_om_table}.

@@

The factorial combinations of fishing histories, observation error
assumptions, and marginal variance and autocorrelation assumptions for
selectivity parameters results in 16 different operating models with
random effects on selectivity (Table \ref{sel_om_table}). Internally,
WHAM treats selectivity parameter \(s\) as a logit-transformed parameter
\[
\log\left(\frac{p_{s,y}-l_{s}}{u_{s}-p_{s,y}}\right) = \mu_s + \epsilon_{s,y}
\] that is a linear combination of a mean \(\mu_s\) and any annual
random effects marginally distributed as
\(\epsilon_{s,y} \sim \text{N}\left(0,\sigma_s^2\right)\) where the
lower and upper bounds of the parameter (\(l_s\) and \(u_s\)) can be
specified by the user. All selectivity parameters are either \(a_50\) or
slope parameters and we assume bounds of 0 and 10 for all selectivity
parameters for all operating and estimating models. The values for the
marginal variance and correlation parameters assumed in the operating
models are given in Table \ref{sel_om_table}.

@@

The factorial combinations of fishing histories, observation error
assumptions, and marginal variance and autocorrelation assumptions for
catchability results in 16 different operating models with random
effects on catchability (Table \ref{q_om_table}). Internally, WHAM
treats catchability for index \(i\) as a logit-transformed parameter \[
\log\left(\frac{q_{i,y}-l_{i}}{u_{i}-q_{i,y}}\right) = \mu_i + \epsilon_{i,y}
\] that is a linear combination of a mean \(\mu_i\) and any annual
random effects marginally distributed as
\(\epsilon_{i,y} \sim \text{N}\left(0,\sigma_i^2\right)\) where the
lower and upper bounds of the catchability (\(l_i\) and \(u_i\)) can be
specified by the user. Here we assume bounds of 0 and 1000 for all
operating and estimating models. For operating and estimation models
with process errors on catchability, the temporal variation is only
assumed for the first index. The values for the marginal variance and
correlation parameters assumed in the operating models are given in
Table \ref{q_om_table}

\hypertarget{estimating-models}{%
\subsection*{Estimating models}\label{estimating-models}}
\addcontentsline{toc}{subsection}{Estimating models}

For each data set simulated from an operating model 20 estimating models
were fit. A total of 32 different estimating models were fit across all
operating models where the subset of 20 depended on the source of
process error (Table \ref{em_table}). The first 20 estimating models in
Table \ref{em_table} were fit to simulate data sets from R and R+S
operating models. Estimating models 5 to 24 in Table \ref{em_table} were
fit to simulate data sets from R+M operating models. Estimating models 5
to 20 and 25-28 in Table \ref{em_table} were fit to simulate data sets
from R+Sel operating models. Finally, estimating models 5 to 20 and
29-32 in Table \ref{em_table} were fit to simulate data sets from R+q
operating models. The observation error variance of aggregate catch and
indices were all assumed known at the true values.

@@

We estimated relative bias \[
\widehat {\text{RB}}\left(\theta\right) = \frac{1}{N} \sum^N_{i = 1} \frac{\widehat \theta_i - \theta_i}{\theta_i}
\] of annual SSB and fully-selected fishing mortality for each
estimation model and constructed 95\% confidence intervals for the
median relative bias using the binomial distribution approach as in
\citet{millerhyun18} and \citet{stockmiller21}. We also estimated bias
of stock-recruit parameters for estimation models that assumed the
Beverton-Holt stock recruit function.

@@

We estimated Mohn's \(\rho\) for SSB, fully-selected fishing mortality,
and recruitment for each estimation model. We estimated 7 peels for each
estimation model. We calculated median 95\% confidence intervals for
Mohn's \(\rho\) using the same methods as that for relative bias.

@@

Marginal AIC most accurately determined the correct source of process
error and correlation structure for R+Sel OMs with low observation error
(Figures \ref{Sel_om_proportion_best_aic_R_MF} to
\ref{Sel_om_proportion_best_aic_SR_ME}). When there was low variability
in selectivity process errors and high observation error, R+q or R+S EMs
were more likely to have the best AIC. Whether mean log-natural morality
or stock recruit relationships were estimated appeared to have little
effect on the performance of marginal AIC. \#\#\# R+q operating models
\{-\}

@@

\hypertarget{aic-performance-for-stock-recruit-relationship}{%
\subsection*{AIC performance for stock-recruit
relationship}\label{aic-performance-for-stock-recruit-relationship}}
\addcontentsline{toc}{subsection}{AIC performance for stock-recruit
relationship}

@@

When mean log-natural mortality is known, all estimating models
regardless of process error assumptions produce unbiased estimates of
SSB over the entire time period for R OMs (Figure
\ref{naa_om_em_R_MF_relbias_ssb}). For R+S OMs, only R+S EMs produce
unbiased estimation of SSB and bias resulting from other EMs increased
with larger variability in survival random effects. Estimating mean
log-natural mortality generally produced greater variability in relative
errors of SSB estimates and a change in sign of bias for EMs assuming
process errors other than R and R+S (Figure
\ref{naa_om_em_R_ME_relbias_ssb}). Estimating a stock-recruit
relationship had little discernable effect on bias of SSB estimation
(Figures \ref{naa_om_em_BH_MF_relbias_ssb}), but in combination with
estimating mean log-natural mortality resulted in large bias of R+S EMs
for R OMs with high observation error, constant fishing pressure and
large variability in recruitment process errors
\ref{naa_om_em_BH_ME_relbias_ssb}).

When mean log-natural mortality is known, SSB estimation bias was small
or non-existent for all EM process error assumptions when the OMs had
low variability of the natural mortality process errors except a
tendency toward negative bias in the terminal years for OMs with high
observation error and a change in fishing pressure over time (Figure
\ref{M_om_em_R_MF_relbias_ssb}). Errors in SSB estimation were most
variable with high variability and autocorrelation of natural mortality
process errors, but confidence intervals generally covered 0 indicating
no evidence of bias. Estimating mean log-natural mortality results in
large variability in SSB relative errors, particularly for OMs with high
observation error and constant fishing pressure and all EMs produced
biased estimation of SSB over all years when OMs had high observation
error, a change in fishing pressure and low variability of natural
mortality process errors (Figure \ref{M_om_em_R_ME_relbias_ssb}). Like R
and R+S OMs, estimating a stock-recruit relationship had little
discernible effect on SSB bias (Figures \ref{M_om_em_BH_MF_relbias_ssb}
and \ref{M_om_em_BH_ME_relbias_ssb})

@@

@@

@@

@@

@@

>>>>

We used the Woods Hole Assessment Model (WHAM) to configure operating
and EMs in our simulation study \citep[]{millerstock20,stockmiller21}.
WHAM is an R package freely available as a github repository. For this
study we used
\href{https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3}{version
1.0.6.9000, commit 77bbd94}. This packages has also been used to
configure operating and EMs for closed loop simulations evaluating
index-based assessment methods \citep{legaultetal23} and is used for
management of haddock, butterfish, American plaice, bluefish, Atlantic
cod, and black sea bass in the Northeast US.

We completed a simulation study with a number of OMs that can be
categorized based on where process error random effects are assumed:
abundance at age (R, R+S), natural mortality (R+M), fleet selectivity
(R+Sel), or index catchability (R+q). For each OM assumptions about
variance of process errors and observations are required and the values
we used were based on a review of the range of estimates from recent
applications of WHAM in management of stocks of haddock, butterfish, and
American plaice in the NE US.

In total, we configured 72 OMs with alternative assumptions about the
source and variability of process errors, level of observation error in
indices and age composition data, and contrast in fishing pressure over
time. We fitted 20 EMs to observations from each of 100 simulations
where process errors were also simulated. EMs made alternative
assumptions about the source of process errors and whether natural
mortality (or the mean for models with process error in natural
mortality) was estimated and whether a Beverton-Holt stock recruit
relationship was estimated within the EM. Details of each of the
operating and EMs are described below.

We did not use the bias-correction feature for process errors or
observations described by \citet{stockmiller21} for operating and EMs.
Simulations and model fitting were all carried out on the University of
Massachusetts Green High-Performance Computing Cluster. All code we used
to perform the simulation study and summarize results can be found at
\url{https://github.com/timjmiller/SSRTWG/tree/main/Project_0/code}.

@@

The population model tracks 10 age classes: ages 1 to 10+ and we assume
spawning occurs 1/4 of the way through the year. The maturity at age was
a logistic curve with \(a_{50}\) = 2.89 and slope = 0.88 and assumed
known in all EMs (Figure \ref{om_maturity}).

@@

We assumed a Beverton-Holt stock recruit function with constant
pre-recruit mortality parameters for all OMs. All post-recruit
productivity components are constant in the NAA and survey catchability
process error OMs. Therefore steepness and unfished recruitment are also
constant over the time period for those OMs \citep{millerbrooks21}. We
specified unfished recruitment = \(R_0 = e^{10}\) and
\(\Fmsy = F_{40\%} = 0.348\) equated to a steepness of 0.69 and
\(\alpha=0.60\) and \(\beta = 2.4 \times 10^{-5}\) for the \[
N_{1,y} = \frac{\alpha \text{SSB}_{y-1}}{1 + \beta \text{SSB}_{y-1}} 
\] Beverton-Holt parameterization (Figure \ref{om_sr}). For OMs without
process errors on natural mortality we assumed the rate was assumed 0.2.
For OMs with process errors on natural mortality the mean rate was 0.2.

Two alternative fishing histories were used for OMs. In the first
scenario, the stock experiences overfishing for the first 20 years and
fishing at \Fmsy for the last 20 years. In the second scenario, the
stock is fished at \Fmsy for the entire time period. The magnitude of
the overfishing assumptions is based on average estimates of overfishing
for NE groundfish stocks from \citet{wiedenmannetal19}.
\Citet{legaultetal23} also used similar approaches to defining fishing
mortality histories for OMs.

@@

The factorial combinations of fishing histories, observation error
assumptions, and marginal variance and autocorrelation assumptions for
natural mortality results in 16 different OMs with random effects on
natural mortality (Table \ref{M_om_table}). Internally, WHAM treats
natural mortality as a log-transformed parameter \[
\log M_y = \mu_M + \epsilon_{M,y}
\] that is a linear combination of a mean \(\mu_M\) and any annual
random effects marginally distributed as
\(\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)\). The values
for the marginal variance and correlation parameters assumed in the OMs
are given in Table \ref{M_om_table}.

@@

The factorial combinations of fishing histories, observation error
assumptions, and marginal variance and autocorrelation assumptions for
selectivity parameters results in 16 different OMs with random effects
on selectivity (Table \ref{sel_om_table}). Internally, WHAM treats
selectivity parameter \(s\) as a logit-transformed parameter \[
\log\left(\frac{p_{s,y}-l_{s}}{u_{s}-p_{s,y}}\right) = \mu_s + \epsilon_{s,y}
\] that is a linear combination of a mean \(\mu_s\) and any annual
random effects marginally distributed as
\(\epsilon_{s,y} \sim \text{N}\left(0,\sigma_s^2\right)\) where the
lower and upper bounds of the parameter (\(l_s\) and \(u_s\)) can be
specified by the user. All selectivity parameters are either \(a_50\) or
slope parameters and we assume bounds of 0 and 10 for all selectivity
parameters for all operating and EMs. The values for the marginal
variance and correlation parameters assumed in the OMs are given in
Table \ref{sel_om_table}.

@@

The factorial combinations of fishing histories, observation error
assumptions, and marginal variance and autocorrelation assumptions for
catchability results in 16 different OMs with random effects on
catchability (Table \ref{q_om_table}). Internally, WHAM treats
catchability for index \(i\) as a logit-transformed parameter \[
\log\left(\frac{q_{i,y}-l_{i}}{u_{i}-q_{i,y}}\right) = \mu_i + \epsilon_{i,y}
\] that is a linear combination of a mean \(\mu_i\) and any annual
random effects marginally distributed as
\(\epsilon_{i,y} \sim \text{N}\left(0,\sigma_i^2\right)\) where the
lower and upper bounds of the catchability (\(l_i\) and \(u_i\)) can be
specified by the user. Here we assume bounds of 0 and 1000 for all
operating and EMs. For operating and EMs with process errors on
catchability, the temporal variation is only assumed for the first
index. The values for the marginal variance and correlation parameters
assumed in the OMs are given in Table \ref{q_om_table}

\hypertarget{estimation-models}{%
\subsection*{Estimation models}\label{estimation-models}}
\addcontentsline{toc}{subsection}{Estimation models}

For each data set simulated from an OM 20 EMs were fit. A total of 32
different EMs were fit across all OMs where the subset of 20 depended on
the source of process error (Table \ref{em_table}). The first 20 EMs in
Table \ref{em_table} were fit to simulate data sets from R and R+S OMs.
EMs 5 to 24 in Table \ref{em_table} were fit to simulate data sets from
R+M OMs. EMs 5 to 20 and 25-28 in Table \ref{em_table} were fit to
simulate data sets from R+Sel OMs. Finally, EMs 5 to 20 and 29-32 in
Table \ref{em_table} were fit to simulate data sets from R+q OMs. The
observation error variance of aggregate catch and indices were all
assumed known at the true values.

@@

We estimated bias as the median of the relative errors across all
simulations for a given OM and EM combination. \[
\text{RE}\left(\theta_i\right) = \frac{\widehat \theta_i - \theta_i}{\theta_i}
\] of annual SSB and fully-selected fishing mortality for each EM and
constructed 95\% confidence intervals for the median relative bias using
the binomial distribution approach as in \citet{millerhyun18} and
\citet{stockmiller21}. We also estimated bias of stock-recruit
parameters for EMs that assumed the Beverton-Holt stock recruit function
and the mean natural mortality rate parameter when estimated.

@@

We estimated Mohn's \(\rho\) for SSB, fully-selected fishing mortality,
and recruitment for each EM. We estimated 7 peels for each EM. We
calculated median 95\% confidence intervals for Mohn's \(\rho\) using
the same methods as that for relative bias.

@@

Marginal AIC most accurately determined the correct source of process
error and correlation structure for R+Sel OMs with low observation error
(Figures \ref{Sel_om_proportion_best_aic_R_MF} to
\ref{Sel_om_proportion_best_aic_SR_ME}). When there was low variability
in selectivity process errors and high observation error, R+q or R+S EMs
were more likely to have the best AIC. Whether mean log-natural morality
or stock recruit relationships were estimated appeared to have little
effect on the performance of marginal AIC.

@@

\hypertarget{aic-performance-for-the-stock-recruit-relationship}{%
\subsection*{AIC performance for the stock-recruit
relationship}\label{aic-performance-for-the-stock-recruit-relationship}}
\addcontentsline{toc}{subsection}{AIC performance for the stock-recruit
relationship}

@@

When mean log-natural mortality is known, all estimating models
regardless of process error assumptions produce unbiased estimates of
SSB over the entire time period for R OMs (Figure
\ref{naa_om_em_R_MF_relbias_ssb}). For R+S OMs, only R+S EMs produce
unbiased estimation of SSB and bias resulting from other EMs increased
with larger variability in survival random effects. Estimating mean
log-natural mortality generally produced greater variability in relative
errors of SSB estimates and a change in sign of bias for EMs assuming
process errors other than R and R+S (Figure
\ref{naa_om_em_R_ME_relbias_ssb}). Estimating a stock-recruit
relationship had little discernable effect on bias of SSB estimation
(Figures \ref{naa_om_em_BH_MF_relbias_ssb}), but in combination with
estimating mean log-natural mortality resulted in large bias of R+S EMs
for R OMs with high observation error, constant fishing pressure and
large variability in recruitment process errors (Figure
\ref{naa_om_em_BH_ME_relbias_ssb}).

When mean log-natural mortality is known, SSB estimation bias was small
or non-existent for all EM process error assumptions when the OMs had
low variability of the natural mortality process errors except a
tendency toward negative bias in the terminal years for OMs with high
observation error and a change in fishing pressure over time (Figure
\ref{M_om_em_R_MF_relbias_ssb}). Errors in SSB estimation were most
variable with high variability and autocorrelation of natural mortality
process errors, but confidence intervals generally covered 0 indicating
no evidence of bias. Estimating mean log-natural mortality results in
large variability in SSB relative errors, particularly for OMs with high
observation error and constant fishing pressure and all EMs produced
biased estimation of SSB over all years when OMs had high observation
error, a change in fishing pressure and low variability of natural
mortality process errors (Figure \ref{M_om_em_R_ME_relbias_ssb}). Like R
and R+S OMs, estimating a stock-recruit relationship had little
discernible effect on SSB bias (Figures \ref{M_om_em_BH_MF_relbias_ssb}
and \ref{M_om_em_BH_ME_relbias_ssb}).

@@

\hypertarget{rq-operating-models-3}{%
\subsubsection*{R+q operating models}\label{rq-operating-models-3}}
\addcontentsline{toc}{subsubsection}{R+q operating models}

@@

\hypertarget{r-rs-operating-models-4}{%
\subsubsection*{R, R+S operating models}\label{r-rs-operating-models-4}}
\addcontentsline{toc}{subsubsection}{R, R+S operating models}

@@

\hypertarget{rm-operating-models-4}{%
\subsubsection*{R+M operating models}\label{rm-operating-models-4}}
\addcontentsline{toc}{subsubsection}{R+M operating models}

@@

\hypertarget{rsel-operating-models-4}{%
\subsubsection*{R+Sel operating models}\label{rsel-operating-models-4}}
\addcontentsline{toc}{subsubsection}{R+Sel operating models}

@@

\hypertarget{rq-operating-models-4}{%
\subsubsection*{R+q operating models}\label{rq-operating-models-4}}
\addcontentsline{toc}{subsubsection}{R+q operating models}
